\Huge \bf \sc Design and implementation of sensor fusion and control for an autonomous quadrotor

### 2013-07-11 latex/paper/paper.tex
terminalo bo

\begin{abstract}
%\boldmath
This paper describes the design and integration of an instrumentation and control system that allows the autonomous flight of a quadrotor. A comercial frame is acquired, a mathematical model for the quadrotor is developed and its parameters determined from the characterization of the unit. An external intelligence is integrated to play the role of the flight controller. A 9 degrees of freedom Inertial Measurement Unit (IMU) equipped with a barometer is calibrated and added to the platform. Data from the IMU is combined with the information provided by a GPS within a Extended Kalman Filter (EKF) to obtain a reliable estimation of the state variables. The control actions are obtained from a proportional-integral controller based on the LQR algorithm. As shown in section \refp{sec:results-tests}, a stable autonomous platform is achieved.%For windless environments, a stable platform is achieved.
% -cambio-
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

A quadrotor platform can be imagined performing countless tasks, such as search and rescue, wild fire suppression, or scientific research\cite{bib:apps}. Many applications are being developed based on such platforms, but how do they fly? An aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining the development of the mathematical model \cite{bib:quadrotor-bible}\cite{bib:curso_quad} to represent the system, the filtering techniques applied for sensor data fusion\cite{bib:kalman}\cite{bib:kalman2} and the control system\cite{bib:quadrotor-bible}\cite{bib:lqrnotes} that allows the quadrotor to fly.
% -cambio-
The platform is based on the commercial radio controlled quadrotor, shown in figure \refp{fig:commercial_quadrotor}. The length between opposite propelers is $61.5\;cm$, the wheight is $990\;g$, and it has $1300\;g$ of payload. The frame and the electronic speed controllers (ESCs) used for the motores were preserved, whereas the IMU and intelligence were replaced by the flight controller that was developed. A BeagleBoard\footnote{BeagleBoard development board - \url{http://beagleboard.org/}} running Linux\footnote{Angstrom distribution: \url{http://www.angstrom-distribution.org/}} performs the computations requiered to convert raw data received from the IMU\footnote{\textit{Mongoose} IMU - \url{http://store.ckdevices.com/}} over a UART and combine it using an EKF. The EKF overcomes the problems inherent to each sensor and filters out noise, providing a reliable estimation of the state vector. Once the current state is known, the LQR algorithm\cite{bib:lqrnotes}\cite{bib:lqr-discreto} is used to derive the control actions required to bring the system to the desired setpoint.
% -cambio-
\begin{figure}[h!]
	\centering
	\includegraphics[width=.6\columnwidth]{./pics_paper/foto_quad.jpg}
	\caption{\textbf{Commercial quadrotor acquired.}}
	\label{fig:commercial_quadrotor}
\end{figure}

@@

A diagram of the quadrotor is shown in figure \refp{fig:quad}. Two of the motors rotate clockwise (2 and 4) and the other two (1 and 3) rotate counterclockwise. This configuration allows the quadrotor to rotate, tilt and gain/lose altitud by setting different speeds on each motor. Two frames of reference (figure \refp{fig:quad}) are constantly used through out this paper: an intertial frame $S_I - \lbrace \hat{i},\hat{j},\hat{k}\rbrace$ ($\lbrace \vec{x},\vec{y},\vec{z}\rbrace$), relative to the Earth, mapped to  North, West and Up respectively, and a non-intertial frame $S_q - \lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$ ($\lbrace \vec{x}_q,\vec{y}_q,\vec{z}_q\rbrace$) relative to the quadrotor. The mapping of one frame to the other can be achieved by applying the three rotations shown in figure \refp{fig:rotaciones}. The angles $\lbrace \theta, \varphi, \psi\rbrace$ are known as Euler angles.

\begin{figure}[h!]
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/quad_coord.jpg}
	\vspace{-5pt}
	\caption{\textbf{Model of the quadrotor -} The blue arrows represent the intertial reference frame $S_I$, and the red arrows represent the non-inertial reference frame $S_q$. The cyan ``looped'' arrows indicate the direction of rotation of each motor, which rotate at $\omega_i$ and generate a torque $M_i$ opposite to their direction of rotation. The arrows labeled $T_{[1,2,3,4]}$ represent the thrust of the motors. The semicircle and the two yellow spheres indicate the $x_q$ axis of the unit.}
    \vspace{-10pt}
	\label{fig:quad}
\end{figure}
\subsection{Dynamics-kinematics of the system}
\label{sec:modelo-dyn-kin}

@@

@@

The IMU is equiped with the following sensors:
\begin{itemize}
\item \textbf{Barometer:} Measures the absolute pressure of the environment. Variations of pressure are used to estimate variations in the altitud of the system.
\item \textbf{Thermometer:} The barometer includes a thermometer. The temperature data is used to apply a temperature compensation to the calibrations performed on the gyroscope and the accelerometer.
\item \textbf{Gyroscope:} A 3-axis gyroscope is used to measure angular velocity of $S_q$. A calibration, based on \cite{bib:calib_imu}, was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Accelerometer:} A 3-axis accelerometer is used to measure gravity. Under the hypothesis that no other accelerations are present, this allows the determination of two of the three Euler angles: $\{\psi, \phi\}$. This hypothesis is acceptable, since the accelerations involved are not significant compared to gravity. A calibration, based on \cite{bib:calib_imu}, was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Magnetometer:} In an area free of magnetic interference this 3-axis sensor will measure $\vec{B}$, the Earth's magnetic field, allowing to determine what direction is North. If the system is horizontal (or the inclination is estimated using other sensors) this sensor can be used to determine the last of the three Euler angles: $\theta$. A calibration based on \cite{bib:bola} and \cite{bib:alain} was performed on this sensor.
\end{itemize}

@@

In order to perform adequate control actions, a reliable estimation of the state variables must be available in real time. The Kalman Filter uses the mathematical model for the system to predict what should happen next given the current state, and corrects the prediction with the information read from the sensors, taking into consideration how much confidence is placed on the prediction and how much on the measurements. This weighted prediction-correction technique allows a smooth state estimation without the typical delay introduced by filtering, even small delays can severly affect the performance of the system.

Every sensor has its issues: The gyro drifts over time; the accelerometer is very sensible to the vibrations generated by the motors; the magnetometer is distorted by ferromagnetic materials; the GPS has a very poor accuracy and a slow update rate. Each sensor by itself is very limited, but they can be combined to compensate for their limitations. The filter takes care of this.

%TODO pasar figura a ingl\'e
\begin{figure}
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_kalman_eng.png}
	\caption{\textbf{EKF - } Outline of how sensor data is combined to estimate the state variables.}
	\label{fig:diagrama_kalman}
\end{figure}

\subsection{EKF}
\label{sec:kalman-ekf}

The theory behind a standard Kalman Filter does not hold for a system that is not linear. The model for the quadrotor given by \refp{eq:modelo} is highly non-linear, so an EKF is implemented based on \cite{bib:kalman} and \cite{bib:kalman2}. An EKF is compatible with non-linear systems, but it is not optimal, the performance is highly dependant on the linearizacion \cite{bib:kay}. The implementation of the EKF gave very good results, and proved to play a critical part in the system. As an example, the data provided by the accelerometer is ``unusable'' without filtering, and experiments with a simple low pass filter (LPF) showed that a 60ms delay introduced by the LPF severly deteriorated the performance of the system. When the EKF was assigned the task of reducing noise, the perfomance was significatively improved.

@@

% \begin{figure}
% 	\centering
% 	\includegraphics[width=1\columnwidth]{./pics_paper/xyz.pdf}
% 	\caption{}
% 	\label{fig:xyz.pdf}
% \end{figure}

@@

Figure \refp{fig:diagrama_kalman} shows a diagram of how data from the sensors is combined within the EKF, assisted by the model of the system. The output of the filter is an estimation the all the variables of the state vector.

@@

The control system is shown in figure (\refp{fig:diagrama_bloques_eng.pdf}). The feedback matrices have many entries, $K_p$ has 48 and $K_i$ has 16. A \textit{root locus} or \textit{pole-zero} analysis on such a system is not a simple task, additional complications are introduced by the fact that $K_p$ and $K_i$ will change when the trayectory changes, they depend on the linearization of the system. A simple and automated solution is achieved by means of the LQR algorithm. The initial design did not include the $K_i$ term, it was introduced to take into consideration errors in the characterization and modelling of the system.

\begin{figure}
\vspace{-60pt}
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_bloques_eng.pdf}
\vspace{-180pt}
	\caption{\textbf{Control System:} $x_{sp}$ and $u_{sp}$ represent the setpoints for the state vector and the speed of each motor. The $K_p$ and $K_i$ blocks are the proportional and integral gain matrices. The output of the system is the current state vector.}
	\label{fig:diagrama_bloques_eng.pdf}
\end{figure}

@@

where $Q$ and $R$ are positive-definite matrices, representing the energy of the controlled states and of the control signal, respectively. The choice of $Q$ and $R$ is a tradeoff between the error that is tolerated and the energy that the system can use to correct it.

@@

The software runs on several independant boards:

@@

The code implements a discrete version of the LQR control system mentioned in \refp{sec:control}, based on a discretization of the model described in \refp{sec:modelo}. There are some practical considerations worth mentioning:
\begin{itemize}
\item \textbf{Saturation:} The integral term saturates it is designed to compensate for small errors in characterization/modelling and should never grow too much. Large diferences between the current state and the setpoint should be handled by the proportional term.
\item \textbf{Limiting:} The control block sets $\vert \vec{X} - \vec{X}_{sp}\vert < \vert \vec{X}_{th} \vert$ to avoid attempts to perform actions that the system cannot handle.
\end{itemize}

\section{Results}

\subsection{Tests}
\label{sec:results-tests}

\begin{figure}[h!]
  \centering
  \subfloat[Roll angle deviation]{\label{fig:psimetro_psi}
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_psi.pdf}} \hspace*{-5pt}
  \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each motor]{\label{fig:psimetro_w} 
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_w.pdf}}
  \caption{\textbf{Stablization experiment -} The Roll angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
  \label{fig:rotaciones}
\end{figure}

The testing phase was separated into several parts:

\begin{enumerate}
\item \textbf{Basic Stability}: During this first stage, the quadcopter was only allowed to move (and correct) one of the Euler angles $\{\psi,\phi\}$. The control system was tuned until an acceptable behavior was obtained. This procedure was applied to each angle. Once this stage was completed the quadcopter was able to mantain horizontality by controlling $\{\psi,\phi,\omega_{q_x},\omega_{q_y}\}$.
\item \textbf{Orientation}: To verify orientation was correctly maintained the speed of the motors was reduced, below the level required to lift off, and the quadcopter was hung from strings. During this test offsets can be set on each motors to compensate for differences between them. After this test, the control system was able to hold the last of the Euler angles: $\{\theta,\omega_{q_z}\}$.
\item \textbf{Limited flying}: At this point the three Euler angles are controlled, so a basic level of stability is expected. During this stage strings were attached to the quadcopter from the top and bottom, and it was allowed to fly. After this test, the control system was able to take-off and to hold altitud and vertical speed: $\{z,v_{q_z}\}$.
\item \textbf{Free flight}: The last step was to let the quadcopter fly. This stage required the GPS. After it was completed, the control system was able to hold it's horizontal position $\{x,y,v_{q_x},v_{q_y}\}$, completing control over the state variables.
\end{enumerate}

Free flight tests were limited due to the poor reliabitily of the ESCs provided by LotusRC\footnote{\url{http://www.lotusrc.com/}}, which showed issues such as overheating until burning out and random glitches that would cause motors to turn off.

\subsection{System response}
\label{sec:results-system-response}

To ilustrate some results, figure \refp{fig:psi_esc.pdf} shows the step response for $\psi$ during an experiment where this was the only degree of freedom. Figure \refp{fig:altura.pdf} shows altitud during takeoff from altitud 0m until a target altitud of 1m.

\begin{figure}
	\centering
	\includegraphics[width=.9\columnwidth]{./pics_paper/control_escalon_eng.png}
	\caption{\textbf{Step response:} Performance during an experiment limited to a single degree of freedom ($\psi$).}
	\label{fig:psi_esc.pdf}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/altura.pdf}
	\caption{\textbf{Altitud:} Performance during takeoff (from 0m to 1m).}
	\label{fig:altura.pdf}
\end{figure}

Flight test were satisfactory in windless environments. In presence of wind the system is not stable. The aluminium protections added (to protect the motors from the faulty ESCs) have flat surfaces that may intensify the effect of the wind.

@@

The main goal of the project was achieved: A quadcopter able to fly on its own. The system was successfully tested. A basic control was implemented, capable of stabilizing the system. The doors are open for future projects to add additional levels of intelligence allowing the quadcopter to perform tasks such as power line inspection, surveillance, assitance during natural disasters, etc.

\subsection{Future Work}
\begin{itemize}
\item The LotusRC ESCs proved to be extremely unreliable\footnote{Several versions were tested, all of then showed the same issues.} and severly delayed the project. Replacing the ESCs is a critical task.
\item The choice of implementing the flight controller on an operating system that does not run in real-time made timing a complicated issue. These issues were dealt with, but assinging additional tasks to the system will surely distrupt timing. Porting the stabilization system, which requires a very fast and steady loop, to a dedicated microprocessor would be a significant improvement. It would allow the code running on top of the operating system to perform higher level tasks, such as Visual Simultaneous Localization and Mapping (VSLAM), navigation, tracking, etc.
\item The fact that the system does not work in a windy environment severly limits the range of possible applications, solving this issue by either reducing the aerodynamic resistance or implementing a different control system would be a significant improvement.
%TODO paper q dijo el pater del viento.
\end{itemize}

@@

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank family and friends for their support during this year, and the staff of the University for their commitment to the project.

>>>>

%\vspace*{-.5cm}
\begin{abstract}
%\boldmath
This paper describes the design and integration of an instrumentation and control system that allows the autonomous flight of a quadrotor. A comercial frame is acquired, a mathematical model for the quadrotor is developed and its parameters determined from the characterization of the unit. An external intelligence is integrated to play the role of the flight controller. A 9 degrees of freedom Inertial Measurement Unit (IMU) equipped with a barometer is calibrated and added to the platform. Data from the IMU is combined with the information provided by a GPS within a Extended Kalman Filter (EKF) to obtain a reliable estimation of the state variables. The control actions are obtained from a proportional-integral controller based on the LQR algorithm. As shown in section \refp{sec:results-tests}, a stable autonomous platform is achieved.%For windless environments, a stable platform is achieved.
% -cambio-
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

A quadrotor platform can be imagined performing countless tasks, such as search and rescue, wild fire suppression, or scientific research\cite{bib:apps}. Many applications are being developed based on such platforms, but how do they fly? An aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining the development of the mathematical model \cite{bib:quadrotor-bible}\cite{bib:curso_quad} to represent the system, the filtering techniques applied for sensor data fusion\cite{bib:kalman}\cite{bib:kalman2} and the control system\cite{bib:quadrotor-bible}\cite{bib:lqrnotes} that allows the quadrotor to fly.
% -cambio-
The platform is based on the commercial radio controlled quadrotor, shown in figure \refp{fig:commercial_quadrotor}. The length between opposite propellers is $61.5\;cm$, the weight is $990\;g$, and it has $1300\;g$ of payload. The frame and the electronic speed controllers (ESCs) used for the motors were preserved, whereas the IMU and intelligence were replaced by the flight controller that was developed. A BeagleBoard\footnote{BeagleBoard development board - \url{http://beagleboard.org/}} running Linux\footnote{Angstrom distribution: \url{http://www.angstrom-distribution.org/}} performs the computations required to convert raw data received from the IMU\footnote{\textit{Mongoose} IMU - \url{http://store.ckdevices.com/}} over a UART and combine it using an EKF. The EKF overcomes the problems inherent to each sensor and filters out noise, providing a reliable estimation of the state vector. Once the current state is known, the LQR algorithm\cite{bib:lqrnotes}\cite{bib:lqr-discreto} is used to derive the control actions required to bring the system to the desired set-point.
% -cambio-
\begin{figure}[h!]
%\vspace{-.8cm}
	\centering
	\includegraphics[width=.8\columnwidth]{./pics_paper/javi.png}
	\caption{\textbf{Commercial quadrotor acquired.}}
	\label{fig:commercial_quadrotor}
\end{figure}

@@

A diagram of the quadrotor is shown in figure \refp{fig:quad}. Two of the motors rotate clockwise (2 and 4) and the other two (1 and 3) rotate counterclockwise. This configuration allows the quadrotor to rotate, tilt and gain/lose altitud by setting different speeds on each motor. Two frames of reference (figure \refp{fig:quad}) are constantly used through out this paper: an intertial frame $S_I - \lbrace \hat{i},\hat{j},\hat{k}\rbrace$ ($\lbrace \vec{x},\vec{y},\vec{z}\rbrace$), relative to the Earth, mapped to  North, West and Up respectively, and a non-inertial frame $S_q - \lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$ ($\lbrace \vec{x}_q,\vec{y}_q,\vec{z}_q\rbrace$) relative to the quadrotor. The mapping of one frame to the other can be achieved by applying the three rotations shown in figure \refp{fig:rotaciones}. The angles $\lbrace \theta, \varphi, \psi\rbrace$ are known as Euler angles.

\subsection{Dynamics-kinematics of the system}
\label{sec:modelo-dyn-kin}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/quad_coord.jpg}
	\vspace{-5pt}
	\caption{\textbf{Model of the quadrotor -} The blue arrows represent the inertial reference frame $S_I$, and the red arrows represent the non-inertial reference frame $S_q$. The cyan ``looped'' arrows indicate the direction of rotation of each motor, which rotate at $\omega_i$ and generate a torque $M_i$ opposite to their direction of rotation. The arrows labeled $T_{[1,2,3,4]}$ represent the thrust of the motors. The semicircle and the two yellow spheres indicate the $x_q$ axis of the unit.}
    %\vspace{-10pt}
	\label{fig:quad}
\end{figure}

@@

The IMU is equiped with the following sensors:
\begin{itemize}
\item \textbf{Barometer:} Measures the absolute pressure of the environment. Variations of pressure are used to estimate variations in the altitude of the system.
\item \textbf{Thermometer:} The barometer includes a thermometer. The temperature data is used to apply a temperature compensation to the calibrations performed on the gyroscope and the accelerometer.
\item \textbf{Gyroscope:} A 3-axis gyroscope is used to measure angular velocity of $S_q$. A calibration, based on \cite{bib:calib_imu}, was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Accelerometer:} A 3-axis accelerometer is used to measure gravity. Under the hypothesis that no other accelerations are present, this allows the determination of two of the three Euler angles: $\{\psi, \phi\}$. This hypothesis is acceptable, since the accelerations involved are not significant compared to gravity. A calibration, based on \cite{bib:calib_imu}, was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Magnetometer:} In an area free of magnetic interference this 3-axis sensor will measure $\vec{B}$, the Earth's magnetic field, allowing to determine what direction is North. If the system is horizontal (or the inclination is estimated using other sensors) this sensor can be used to determine the last of the three Euler angles: $\theta$. A calibration based on \cite{bib:bola} and \cite{bib:alain} was performed on this sensor.
\end{itemize}

@@

In order to perform adequate control actions, a reliable estimation of the state variables must be available in real time. The Kalman Filter uses the mathematical model for the system to predict what should happen next given the current state, and corrects the prediction with the information read from the sensors, taking into consideration how much confidence is placed on the prediction and how much on the measurements. This weighted prediction-correction technique allows a smooth state estimation without the typical delay introduced by filtering, even small delays can severely affect the performance of the system.

Every sensor has its issues: The gyro drifts over time; the accelerometer is very sensible to the vibrations generated by the motors; the magnetometer is distorted by ferromagnetic materials; the GPS has a very poor accuracy and a slow update rate. Each sensor by itself is very limited, but they can be combined to compensate for their limitations. The filter takes care of this by integrating all sensors in order to obtain a more accurate state estimation. Figure \refp{fig:diagrama_kalman} shows a diagram of how data from the sensors is combined within the filter (EKF), assisted by the model of the system.
\begin{figure}[h!]
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_kalman_eng.png}
	\caption{\textbf{EKF - } Outline of how sensor data is combined to estimate the state variables.}
	\label{fig:diagrama_kalman}
\end{figure}
Euler angles are primary estimated from the combination of Magnetometer and Accelerometer measurements and used as one of the filter inputs. From that estimation of Euler angles and the Accelerometer measurements can be deduced quadrotor velocities $\{v_{qx},\;v_{qy}\;v_{qz}\}$ by subtracting gravity and an integration, which are also introduced to the filter. Angular velocities referenced to the quadrotor and height can be easily deduced from Gyroscope and Barometer measurements respectively, which are also introduced to the filter. Furthermore the GPS measurements can be also used as filter input. All this primary estimations from sensors are considered as measurements in the Kalman Filter, while the prediction is based, as said, in the dynamic model. The output of the filter is an estimation the all the variables of the state vector.

\subsection{Extended Kalman Filter}
\label{sec:kalman-ekf}

The theory behind a standard Kalman Filter does not hold for a nonlinear system. The model for the quadrotor given by \refp{eq:modelo} is highly nonlinear, so an Extended Kalman Filter (EKF) is implemented based on \cite{bib:kalman} and \cite{bib:kalman2}. While the Kalman Filter ensures a statistical optimal performance, the EKF is not optimal, and it is not possible to determine the error a priori, due to high dependency of the performance with the linearization \cite{bib:kay}. Although, EKF is the most used and popular filtering technique in navigation problems. The implementation of the EKF gave very good results, and proved to play a critical part in the system. As an example, the data provided by the accelerometer is ``unusable'' without filtering, and experiments with a simple low pass filter (LPF) showed that a 60ms delay introduced by the LPF severely deteriorated the performance of the system. When the EKF was assigned the task of reducing noise, the performance was significantly improved.

 \begin{figure}
	\hspace{-.8cm}
	\includegraphics[width=1.12\columnwidth]{./pics_paper/kalman.pdf}
	\caption{\textbf{Kalman -} State estimation of the three Euler angles. In each graphic is shown the sensors estimation of each angle and the corresponding EKF estimation.}
	\label{fig:kalman}
\end{figure}

In each graphic of figure\refp{fig:kalman} is shown the primary estimation obtained from sensors and the kalman estimation of that state variable for the three Euler angles. As can be seen, the noise is greatly reduced and no delay is introduced.

@@

@@

@@

The control system is shown in figure (\refp{fig:diagrama_bloques_eng.pdf}). The feedback matrices have many entries, $K_p$ has 48 and $K_i$ has 16. A \textit{root locus} or \textit{pole-zero} analysis on such a system is not a simple task, additional complications are introduced by the fact that $K_p$ and $K_i$ will change when the trajectory changes, they depend on the linearization of the system. A simple and automated solution is achieved by means of the LQR algorithm. The initial design did not include the $K_i$ term, it was introduced to take into consideration errors in the characterization and modeling of the system.

\begin{figure}
%	\vspace{-60pt}
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_bloques_eng.png}
%	\vspace{-180pt}
	\caption{\textbf{Control System:} $x_{sp}$ and $u_{sp}$ represent the set-points for the state vector and the speed of each motor. The $K_p$ and $K_i$ blocks are the proportional and integral gain matrices. The output of the system is the current state vector.}
	\label{fig:diagrama_bloques_eng.pdf}
\end{figure}

@@

where $Q$ and $R$ are positive-definite matrices, representing the energy of the controlled states and of the control signal, respectively. The choice of $Q$ and $R$ is a trade-off between the error that is tolerated and the energy that the system can use to correct it.

@@

The software runs on several independent boards:

@@

The code implements a discrete version of the LQR control system mentioned in \refp{sec:control}, based on a discretization of the model described in \refp{sec:modelo}. There are some practical considerations worth mentioning:
\begin{itemize}
\item \textbf{Saturation:} The integral term saturates it is designed to compensate for small errors in characterization/modelling and should never grow too much. Large differences between the current state and the set-point should be handled by the proportional term.
\item \textbf{Limiting:} The control block sets $\vert \vec{X} - \vec{X}_{sp}\vert < \vert \vec{X}_{th} \vert$ to avoid attempts to perform actions that the system cannot handle.
\end{itemize}

\section{Results}
\label{sec:results-tests}

\subsection{Basic stability}

During this first stage, the quadrotor was only allowed to move (and correct) one of the following Euler angles: $\{\psi,\phi\}$. The control system was tuned until an acceptable behavior was obtained. This procedure was applied to each angle. Once this stage was completed the quadrotor was able to maintain horizontality by controlling $\{\psi,\phi,\omega_{q_x},\omega_{q_y}\}$. The behavior for the Roll ($\psi$) angle is shown in figure \refp{fig:psimetro}. While in figure \refp{fig:psimetro_psi} is shown the perturbation imposed to the \emph{Roll} angle, the reaction can be observed in figure \refp{fig:psimetro_w}. The system step response for a step of more than $20^o$ shows a overshoot of $3^o$ or $4^o$ and a rise time of about $0.4\;s$, which means an excellent performance, as can be seen in figure \refp{fig:psimetro_psi}.
\begin{figure}[h!]
  \centering
  \subfloat[Roll angle deviation]{\label{fig:psimetro_psi}
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_psi.pdf}} \hspace*{-5pt}
  \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each motor]{\label{fig:psimetro_w} 
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_w.pdf}}
  \caption{\textbf{Stablization experiment -} The Roll angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
  \label{fig:psimetro}
\end{figure}

\subsection{Orientation}
To verify orientation was correctly maintained, the speed of the motors was reduced below the level required to lift off, and the quadrotor was hung from strings. After this test, the control system was able to hold the last of the Euler angles: $\{\theta\}$. 
\begin{figure}[h!]
  \centering
  \subfloat[Yaw angle deviation]{\label{fig:theta}
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/theta.pdf}} \hspace*{-5pt}
  \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each pair of motors.]{\label{fig:theta_w} 
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/theta_w.pdf}}
  \caption{\textbf{Orientation experiment -} The Yaw angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
  \label{fig:theta_all}
\end{figure}
In the figure \refp{fig:theta_all} is shown the system behavior against a perturbation in yaw angle. To rotate, the system must generate a torque in the opposite direction of the perturbation by unbalancing the angular velocity of the motors in pairs: $\omega_1$ and $\omega_3$ in one direction and $\omega_2$ and $\omega_4$ in the other. In this case, the rise time is of about $1.6\;s$, much higher than in the other two Euler angles, which can be permitted as Yaw angle is not such critical for stability. The overshoot obtained is of about 13\% of the step size.

\subsection{Limited Flying}
\label{sec:results-system-response}
At this point the three Euler angles are controlled, so a basic level of stability is expected, and the quadrotor is allowed to fly. After this test, the control system was able to take-off and to hold altitud and vertical speed: $\{z,v_{q_z}\}$.

%\textbf{Free flight}: The last step was to let the quadrotor fly. This stage required the GPS. After it was completed, the control system was able to hold it's horizontal position $\{x,y,v_{q_x},v_{q_y}\}$, completing control over the state variables.

%Free flight tests were limited due to the poor reliabitily of the ESCs provided by LotusRC\footnote{\url{http://www.lotusrc.com/}}, which showed issues such as overheating until burning out and random glitches that would cause motors to turn off.

%To ilustrate some results, figure \refp{fig:psi_esc.pdf} shows the step response for $\psi$ during an experiment where this was the only degree of freedom.

Figure \refp{fig:altura.pdf} shows altitude during takeoff from altitude 0m until a target altitude of 1m is achieved. As can be seen, the system presents an overshoot of about a meter, which is considered as acceptable for many applications.

%\begin{figure}
%	\centering
%	\includegraphics[width=.8\columnwidth]{./pics_paper/control_escalon_eng.png}
%	\caption{\textbf{Step response:} Performance during an experiment limited to a single degree of freedom ($\psi$).}
%	\label{fig:psi_esc.pdf}
%\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/altura.pdf}
	\caption{\textbf{Altitud:} Performance during takeoff (from 0m to 1m).}
	\label{fig:altura.pdf}
\end{figure}

%Flight test were satisfactory in windless environments. In presence of wind the system is not stable. The aluminium protections added (to protect the motors from the faulty ESCs) have flat surfaces that may intensify the effect of the wind.

@@

The two main goals of the project were successfuly achieved. First, the design and implementation of the sensor fusion was successfully achieved by the integration of all the measurements in an Extended Kalman Filter which is used for denoising purposes without introducing delay. And then a basic control was implemented, based on LQR algorithm, capable of stabilizing the system. A simplified dynamic model was derived as a first step for both Kalman Filter and controller. \\

The Extended Kalman Filter implemented gave very good results for all the state variables except for the speed of the quadrotor. The main reason is that a speed sensor is not aveilable and the speed can only be infered by integrating the accelerations. Thus, an extra sensor should be used (i.e. a more accurate GPS) or a more sophisticated algorithm should be develped. \\

The controller developed was succesfully tested for the attitude and the height. 
Control: No usa las velocidades Se implemente un controlador que permite hacer muchas cosas, pero solo se testeo el hover y controlando solo algunas cosas. No nos preocupamso por la posici´on horizontal.

The doors are open for future projects to add additional levels of intelligence allowing the quadrotor to perform tasks such as power line inspection, surveillance, assistance during natural disasters, etc.

\begin{figure}
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/tito.png}
	\caption{\textbf{Final setup -} quadrotor fully equipped whith aluminum protections}
	\label{fig:tito}
\end{figure}

The final plataform implemented is shown in figure \refp{fig:tito}. An acrylic box was added in the bottom of the quadrotor to hold the added intelligence. The GPS is attached at the upper part of the quadrotor, so as it has the maximum of satellite visibility. Aluminum protections were added around all motors and propellers so as damages can be prevented.

\subsection{Future Work}
\begin{itemize}
\item The LotusRC ESCs proved to be extremely unreliable\footnote{Several versions were tested, all of then showed the same issues.} and severely delayed the project. Replacing the ESCs is a critical task.
\item The choice of implementing the flight controller on an operating system that does not run in real-time made timing a complicated issue. These issues were dealt with, but assigning additional tasks to the system will surely disrupt timing. Porting the stabilization system, which requires a very fast and steady loop, to a dedicated microprocessor would be a significant improvement. It would allow the code running on top of the operating system to perform higher level tasks, such as Visual Simultaneous Localization and Mapping (VSLAM), navigation, tracking, etc.
%\item The fact that the system does not work in a windy environment severly limits the range of possible applications, solving this issue by either reducing the aerodynamic resistance or implementing a different control system would be a significant improvement.
%TODO paper q dijo el pater del viento.
\end{itemize}

@@

% use section* for acknowledgement
%\section*{Acknowledgment}

%The authors would like to thank family and friends for their support during this year, and the staff of the University for their commitment to the project.
